### 2020-07-16 paper.tex
Checkpoint

\textbf{Notation for dictionaries.} Given a dictionary $x = (A, \mathbf{x})$, we write $A_m$ for $A$ and $m[a]$ for the value $x_a$ associated with a key $a \in A$.
We also denote specific dictionaries using notation $\{k_1 \mapsto v_1, k_2 \mapsto v_2, \ldots\}$.
For example, the dictionary $(A, \mathbf{x})$ with $A = \{1, \mathtt{"foo"}\}$ and $x_1 = 0.123$ and $x_{\mathtt{"foo"}} = 5$ is denoted 
$\{1 \mapsto 0.123, \mathtt{"foo"} \mapsto 5\}$.

@@

\textit{Gaussian mixture model with unknown number of components.} Consider a generative model of univariate data points $y_1, \dots, y_n$ from a Gaussian mixture with an unknown number of components $k$, each with unknown mean $m_i$ and standard deviation $s_i$. If we place a Gaussian prior on $m_i$, a Gamma prior on $s_i$, and a Poisson prior on $k$, the resulting density is
%
$$p(d) = TODO$$
%
when $A_d$ is of the form $\{\texttt{k}, \texttt{y}_\texttt{1}, \dots, \texttt{y}_\texttt{n}\} \cup \{\texttt{m}_i \mid 1 \leq i \leq d[\texttt{k}]\} \cup \{\texttt{s}_i \mid 1 \leq i \leq d[\texttt{k}]\}$, and 0 otherwise. In this case, we have $V_\texttt{k} = \mathbb{N}$ with the counting measure for $\mu_\texttt{k}$, and for all other addresses $a \in \mathcal{A}$, $V_a = \mathbb{R}$ with the Lebesgue measure for $\mu_a$.

@@

Provided the two probabilistic programs $P$ and $Q$ satisfy a simple technical requirement (described in Section~\ref{sec:technical-requirement}), they can be used to specify the model density $p$ and auxiliary densities $p_x$, respectively for use with involutive MCMC on dictionaries (Algorithm~\ref{alg:involutive-mcmc}).
The acceptance probability of involutive MCMC includes the following ratio of densities:
\begin{equation}
\frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)}
\end{equation}
To evaluate this density ratio, we can obtain four traces, one for each dictionary, and use those traces to compute the four probability densities, as shown in Algorithm~\ref{alg:trace-based-involutive-mcmc}.
Section~\ref{sec:sparsity} describes a more efficent approach that exploits sparsity in the involution and resulting cancellation in the ratio for improved efficiency, but requires a more sophisticated probabilistic programming runtime system.

@@

>>>>

\textbf{Notation for dictionaries.} Given a dictionary $d = (A, \mathbf{x})$, we write $A_d$ for $A$ and $d[a]$ for the value $x_a$ associated with a key $a \in A$.
We also denote specific dictionaries using notation $\{k_1 \mapsto v_1, k_2 \mapsto v_2, \ldots\}$.
For example, the dictionary $(A, \mathbf{x})$ with $A = \{1, \mathtt{"foo"}\}$ and $x_1 = 0.123$ and $x_{\mathtt{"foo"}} = 5$ is denoted 
$\{1 \mapsto 0.123, \mathtt{"foo"} \mapsto 5\}$.

@@

\textit{Gaussian mixture model with unknown number of components.} Consider a generative model of univariate data points $y_1, \dots, y_n$ from a Gaussian mixture with an unknown number of components $k$, each with unknown mean $m_i$ and variance $s_i$. If we place a Gaussian prior on $m_i$, an inverse Gamma prior on $s_i$, and a Poisson prior on $k$, the resulting density is
%
\begin{align*}
p(d) =
\begin{array}{l}
p_{\mathrm{poisson}(3)}(d[\mathtt{k}]) \cdot\\
\prod_{i=1}^{d[\mathtt{k}]} p_{\mathrm{normal}(0, 1)}(d[\mathtt{m}_i]) \cdot\\
\prod_{i=1}^{d[\mathtt{k}]} p_{\mathrm{inversegamma}(1, 10)}(d[\mathtt{s}_i]) \cdot\\
\prod_{i=1}^n  \frac{1}{d[\mathtt{k}]} \sum_{j=1}^{d[\mathtt{k}]} \cdot p_{\mathrm{normal}(d[\mathtt{m}_j], d[\mathtt{s}_i])}(d[\mathtt{y}_i])
\end{array}
\end{align*}
%
when $A_d$ is of the form $\{\texttt{k}, \texttt{y}_\texttt{1}, \dots, \texttt{y}_\texttt{n}\} \cup \{\texttt{m}_i \mid 1 \leq i \leq d[\texttt{k}]\} \cup \{\texttt{s}_i \mid 1 \leq i \leq d[\texttt{k}]\}$, and 0 otherwise. In this case, we have $V_\texttt{k} = \mathbb{N}$ with the counting measure for $\mu_\texttt{k}$, and for all other addresses $a \in \mathcal{A}$, $V_a = \mathbb{R}$ with the Lebesgue measure for $\mu_a$.

@@

Provided the two probabilistic programs $P$ and $Q$ satisfy a simple technical requirement (described in Section~\ref{sec:technical-requirement}), they can be used to specify the model density $p$ and auxiliary densities $q_x$ for use with involutive MCMC on dictionaries (Algorithm~\ref{alg:involutive-mcmc}).
The acceptance probability of involutive MCMC includes the following ratio of densities:
\begin{equation}
\frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)}
\end{equation}
This ratio can be computed term-by-term using \textsc{trace-and-score} (Algorithm~\ref{alg:trace-based-involutive-mcmc}). 
Section~\ref{sec:sparsity} describes a more efficient approach that exploits sparsity in the involution and resulting cancellation in the ratio for improved efficiency, but requires a more sophisticated probabilistic programming runtime system.

@@

\textbf{Dynamic checks during inference.} These dynamic assertions can also be run during inference, at each application of the transition kernel. This can be useful to catch bugs that only occur in regions of the state space with low prior mass (but perhaps high posterior mass). When enabled, we can run dynamic checks after each application of the kernel, and when they fail, write to a debugging log and reject the proposed new state. As it turns out, the kernel induced by this procedure is still stationary for $p$:

\begin{lemma}
Let $p$ and $q_x$ be model and auxiliary densities as above, but suppose $f : \mathcal{D} \times \mathcal{D} \rightarrow \mathcal{D} \times \mathcal{D}$ may not be an involution on $Z$.  Trace-based involutive MCMC with dynamic checks enabled, rejecting whenever such a check fails, still yields a kernel that is stationary for $p$.
\end{lemma}
\begin{proof}
Let $R = \{x \in Z \mid f(f(x)) = x \wedge \pi(f(x)) > 0\}$, and let $h(x) := \mathbf{1}[x \in R]f(x) + \mathbf{1}[x \not\in R]x$. Then $h$ is an involution on $Z$, and trace-based involutive MCMC with $p$, $q_x$, and $h$ yields a stationary kernel. But this kernel is the same one induced by using $f$ with dynamic checks. For $x$ on which dynamic checks succeed, $h$ is equivalent to $f$. For $x$ on which dynamic checks fail, $h$ is equivalent to the identity; thus, accepting a move produced by $h$ is equivalent to rejecting.  
\end{proof}
